---
title: "Cell calling"
output:
  html_document: default
  pdf_document: default
---

The following code shows application of four cell calling algorithms (EmptyNN, CellRanger 2.0, EmptyDrops, CellBender) to four datasets: (1) Cell hashing dataset, (2) Multiplexed PBMC dataset, (3) PBMC 8k dataset, and (4) Neuron 900 dataset. See details in our [EmptyNN manuscript](https://www.biorxiv.org/content/10.1101/2021.01.15.426387v1).

# Please install required packages and download datasets before running this analysis (run download_datasets.sh in terminal).
```{r}
## install EmptyNN
# install.packages("devtools")
# library(devtools)
# install_github("lkmklsmn/empty_nn")
## install DropletUtils, a package to run EmptyDrops
#BiocManager::install("DropletUtils")
```


## Load libraries
```{r}
library('EmptyNN')
library('DropletUtils') 
# load other packages
library("Seurat")
library("Matrix")
# R version 3.6.3, EmptyNN 1.0, DropletUtils 1.6.1, Seurat 3.2.3, Matrix 1.3-2
```

# Cell hashing dataset
```{r}
# this file contains raw data matrix and ground-truth label for each barcode.
load("./../data/cell_hashing_raw.RData")
```

## We apply EmptyNN to do the preprocessing.
The **input** is raw count matrix, either in h5 or mtx format. The rows are barcodes and columns are genes.
The **output** is a boolean vector showing cell-free droplet (empty droplet) or cell-containing droplet.
**Runtime** depends on the dataset to be processed and parameter setting. A higher k_folds or iteration means longer runtime. It takes ~30 mins to run with the default parameters.
For demonstration purposes, we set parameters to a low number, which takes ~5 mins.
```{r}
nn.res <- emptynn(Stoeckius.counts,threshold=50,k=2,iteration=1)
```
The **nn.res** is a list, containing 
1. A boolean vector showing the predictions for cell-free or cell-containing droplets
```{r}
nn.res$nn.keep[1:5]
```
2. Probabilities for each barcode in set U
```{r}
head(nn.res$prediction)
```
## Next, we applied other state-of-art methods to benchmark our analysis, including CellRanger 2.0, EmptyDrops, and CellBender.
```{r}
# CellRanger 2.0¶
n_counts <- rowSums(Stoeckius.counts)
ranger.keep <- n_counts>=200
# EmptyDrops
e.out <- emptyDrops(t(Stoeckius.counts))
e.keep <- e.out$FDR <= 0.001
# CellBender is run in a CPU server using following code:
# cellbender remove-background \
#   --input /path/to/Cell_hashing/ \
#   --output /path/to/Cell_hashing_CellBender.h5 \
#   --expected-cells 20000 \
#   --total-droplets-included 25000
```

# Multiplexed PBMC dataset
```{r}
counts <- Read10X_h5("./../data/multiplexed_PBMC_raw.h5")
# EmptyNN
nn.res <- emptynn(t(counts),threshold=100,k=2,iteration=1)
# CellRanger 2.0¶
n_counts <- colSums(counts)
ranger.keep <- n_counts>=200
# EmptyDrops
e.out <- emptyDrops(counts)
e.keep <- e.out$FDR <= 0.001
# CellBender is run in a CPU server using following code:
# cellbender remove-background \
#   --input /path/to/multiplexed_PBMC/ \
#   --output ./multiplexed_PBMC_CellBender.h5 \
#   --total-droplets-included 50000 \
#   --low-count-threshold 50
```

# PBMC 8k dataset
```{r}
# counts <- Read10X_h5("./../data/pbmc_8k_raw.h5")
# # EmptyNN
# nn.res <- emptynn(t(counts),threshold=100,k=2,iteration=1)
# # CellRanger 2.0¶
# n_counts <- colSums(counts)
# ranger.keep <- n_counts>=200
# # EmptyDrops
# e.out <- emptyDrops(counts)
# e.keep <- e.out$FDR <= 0.001
# CellBender is run in a CPU server using following code:
# cellbender remove-background \
#   --input /path/to/multiplexed_PBMC/ \
#   --output ./multiplexed_PBMC_CellBender.h5 \
#   --expected-cells 10000 \
#   --total-droplets-included 25000
```

# Neuron 900 dataset
```{r}
# counts <- Read10X_h5("./../data/neurons_900_raw.h5")
# # EmptyNN
# nn.res <- emptynn(t(counts),threshold=100,k=2,iteration=1)
# # CellRanger 2.0¶
# n_counts <- colSums(counts)
# ranger.keep <- n_counts>=200
# # EmptyDrops
# e.out <- emptyDrops(counts)
# e.keep <- e.out$FDR <= 0.001
# CellBender is run in a CPU server using following code:
# cellbender remove-background \
#   --input /path/to/multiplexed_PBMC/ \
#   --output ./multiplexed_PBMC_CellBender.h5 \
#   --expected-cells 900 \
#   --total-droplets-included 10000
```


