---
title: "Figure 2"
output:
  html_document: default
  pdf_document: default
---
The following code reproduces the Figure 2 in our [EmptyNN manuscript](https://www.biorxiv.org/content/10.1101/2021.01.15.426387v1).

## Please download datasets and seurat objects before running this analysis (run download_datasets.sh in terminal)

## Load libraries
```{r include=FALSE}
library("Seurat")
library("Matrix")
library('pROC')
library("ggplot2")
library("plyr")
library("pheatmap")
load("./../data/BlueYellowColormaps_V1.RData")
# R version 3.6.3, Seurat_3.2.3, Matrix_1.3-2, ggplot_2_3.3.3, pheatmap_1.0.12 
```

## Load (1) raw data
## (2) filtering results of four cell-calling algorithms: EmptyNN (nn.res), CellRanger 2.0 (ranger.keep), EmptyDrops (e.out, e.keep), CellBender (bender.keep)
```{r include=TRUE}
load("./../data/cell_hashing_raw.RData")
load("./../data/cell_hashing_results.RData")
```

# Figure 2A
```{r include=TRUE}
total_counts <- Matrix::rowSums(Stoeckius.counts)
aframe <- data.frame(total_counts, rank = rank(-total_counts), 
                     emptynn = nn.res$nn.keep)
twohundred <- unique(aframe$rank[which(aframe$total_counts == 200)])
fifty <- unique(aframe$rank[which(aframe$total_counts == 50)])
recovered <- aframe$rank[which(aframe$emptynn & 
                                       aframe$total_counts < 200 
                               & aframe$total_counts > 50)]
ggplot(aframe, aes(rank, total_counts)) +
  scale_x_continuous(trans='log10') + scale_y_continuous(trans='log10') +
  geom_point() + theme_bw() +
  xlab("Sorted barcodes") + ylab("Total counts [log10]") +
  geom_vline(xintercept = twohundred, color = "dodgerblue", linetype = "dashed") +
  geom_vline(xintercept = fifty, color = "forestgreen", linetype = "dashed")
```

# Figure 2B
```{r include=TRUE}
ggplot(aframe[which(aframe$total_counts < 250 & aframe$total_counts > 50),], aes(rank,total_counts)) + 
  xlab("Sorted barcodes") + ylab("Total counts") +
  geom_vline(xintercept = twohundred, color = "dodgerblue", linetype = "dashed") +
  geom_vline(xintercept = fifty, color = "forestgreen", linetype = "dashed") +
  geom_vline(xintercept = recovered, color = "darkred", alpha = 0.5) +
  geom_point() + theme_bw()
```

# Figure 2C
```{r include=TRUE}
original_paper <- readRDS("./../data/cell_hashing_original_paper.rds")
genes.use <- VariableFeatures(original_paper)
names(nn.res$nn.keep) <- rownames(Stoeckius.counts)
recover.bcs <- intersect(names(total_counts[total_counts<200]),
                         names(nn.res$nn.keep[nn.res$nn.keep]))
recover <- CreateSeuratObject(counts = t(Stoeckius.counts[recover.bcs,]))
recover <- NormalizeData(recover,verbose = FALSE)
recover <- ScaleData(recover, features = genes.use,verbose = FALSE)
sub.genes.use <- rownames(recover[["RNA"]]@scale.data)
recover <- RunPCA(recover,features=sub.genes.use,verbose = FALSE)
recover <- FindNeighbors(recover, dims = 1:10,verbose = FALSE)
recover <- FindClusters(recover, resolution = 0.2,verbose = FALSE)
recover <- RunTSNE(recover, dims = 1:10, check_duplicates = FALSE,verbose = FALSE)
new.cluster <- c("CD14 Mono","CD4","NK","B",'Platelet')
names(new.cluster) <- levels(recover)
recover <- RenameIdents(recover,new.cluster)
DimPlot(recover,label=T)+NoLegend()+labs(title="Cell type identities of recovered cells")
```

# Figure 2D
```{r}
des <- FindAllMarkers(recover, only.pos = TRUE, min.pct = 0.25,
                      logfc.threshold = 0.25,verbose=FALSE)
asplit_genes <- split(1:nrow(des), des$cluster)
genes <- unlist(lapply(asplit_genes, function(x) des[x[1:5], "gene"]))
genes <- genes[genes %in% rownames(recover@assays$RNA@data)]
# Average cells within each cluster
asplit_cells <- split(rownames(recover@meta.data), recover@active.ident)
means <- do.call(cbind, lapply(asplit_cells, function(x){
  s1 <- Matrix::rowMeans(recover@assays$RNA@data[genes, sample(unlist(x), 10)])
  s2 <- Matrix::rowMeans(recover@assays$RNA@data[genes, sample(unlist(x), 10)])
  s3 <- Matrix::rowMeans(recover@assays$RNA@data[genes, sample(unlist(x), 10)])
  cbind(s1, s2, s3)
}))
cell_type <- unlist(lapply(names(asplit_cells), function(x) rep(x, 3)))
# Create heatmap (sample 3 "replicates")
anno_col <- data.frame(cell_type)
rownames(anno_col) <- colnames(means) <- paste(colnames(means), cell_type)
pheatmap(means,cluster_rows = F, cluster_cols = F, scale = "row",show_rownames = F,
         breaks = seq(-2, 2, length = length(yellow2blue) + 1), col = yellow2blue,
         annotation_col = anno_col,show_colnames = F)
```

# Figure 2E
```{r}
pbmc <- readRDS("./../data/reference_pbmc_3k.rds")
pla <- subset(pbmc,idents="Platelet")
pla.recover <- subset(recover,idents="Platelet")
c1 <- as.matrix(pla.recover[['RNA']]@counts)
c2 <- as.matrix(pla[['RNA']]@counts)
features <- head(rownames(FindMarkers(recover, ident.1 = "Platelet",only.pos = TRUE)),300)
features <- intersect(intersect(rownames(c1),rownames(c2)),features)
expr <- cbind(c1[features,],c2[features,])
m1 <- rowMeans(c1[rownames(expr),])
m2 <- rowMeans(c2[rownames(expr),])
df <- data.frame("recovered"=m1,"bona_fide"=m2)
ggplot(df, aes(x=recovered, y=bona_fide)) + 
        geom_point()+
        geom_smooth(method=lm)+
        xlab("recovered platelets")+ylab("bona fide platelets")
```

# Figure S3A
```{r}
retained <- readRDS("./../data/cell_hashing_retained.rds")
retained$label <- label[colnames(retained),1]
DimPlot(retained,reduction='tsne',group.by = 'label')+
  NoLegend()+labs(title="Cell hashing info")
```
# Figure S3B
```{r}
DimPlot(retained,label=T,reduction='tsne')+NoLegend()+labs(title="Cell type identity")
```

# Figure S3C,D,E,F
```{r}
DimPlot(retained,reduction='tsne',group.by = 'emptynn',cols=c('grey','steelblue'))+
  NoLegend()+labs(title="EmptyNN")
DimPlot(retained,reduction='tsne',group.by = 'cellranger',cols=c('grey','steelblue'))+
  NoLegend()+labs(title="CellRanger 2.0")
DimPlot(retained,reduction='tsne',group.by = 'emptydrops',cols=c('grey','steelblue'))+
  NoLegend()+labs(title="EmptyDrops")
DimPlot(retained,reduction='tsne',group.by = 'cellbender',cols=c('grey','steelblue'))+
  NoLegend()+labs(title="CellBender")
```

# Figure 2F
```{r}
sub <- retained[,sample(colnames(retained),10000)]
sub <- subset(sub,idents = c("Ambient","Doublets","pDCs"), invert=TRUE)
dist.mtx <- as.matrix(dist(sub@reductions$pca@cell.embeddings,
                           method = "euclidean",upper=TRUE))
sub$source <- 'recovered'
sub$source[rownames(sub@meta.data) %in% colnames(original_paper)] <- 'original'
sub$label <- Idents(sub)
# define function euclidean.calc()
euclidean.calc <- function(cluster){
  # recovered with original
  condition <- sub$label==cluster & sub$emptynn & sub$source == 'recovered'
  row <- rownames(sub@meta.data[condition,])
  column <- names(sub$source[sub$source=='original'])
  df <- dist.mtx[row,column]
  neg.m <- data.frame("dist"=colMeans(df))
  idx <- match(rownames(neg.m),names(sub$label))
  neg.m$label <- sub$label[idx]
  colnames(neg.m) <- c("dist",'label')
  # original with original
  condition <- sub$label==cluster & sub$source == 'original'
  column <- rownames(sub@meta.data[condition,])
  dist.b <- dist.mtx[column,column]
  orig.m <- data.frame('dist'=colMeans(dist.b),cluster)
  colnames(orig.m) <- c("dist",'label')
  # recover with recover
  condition <- sub$label==cluster & sub$emptynn & sub$source == 'recovered'
  column <- rownames(sub@meta.data[condition,])
  dist.b <- dist.mtx[column,column]
  recov.m <- data.frame('dist'=colMeans(dist.b),cluster)
  colnames(recov.m) <- c("dist",'label')
  tmp <- rbind(neg.m,orig.m,recov.m)
  tmp$group <- c(rep("recovered with original",nrow(neg.m)),
                 rep("original with original",nrow(orig.m)),
                 rep("recover with recover",nrow(recov.m)))
  return(tmp)
}
# calculate for NK cell type
tmp <- euclidean.calc("NK")
ggplot(tmp, aes(x=label, y=dist,fill=group)) + 
  geom_boxplot(outlier.shape=NA)+
  xlab("original clusters")+ylab("Euclidean distance")
```

# Figure 2G
```{r}
tmp <- retained[,retained$emptynn==TRUE]
m1 <- FindMarkers(tmp, ident.1 = "CD4",ident.2 = "CD14 Mono",only.pos = TRUE,verbose=F)
tmp <- retained[,retained$cellranger==TRUE]
m2 <- FindMarkers(tmp, ident.1 = "CD4",ident.2 = "CD14 Mono",only.pos = TRUE,verbose=F)
tmp <- retained[,retained$emptydrops==TRUE]
m3 <- FindMarkers(tmp, ident.1 = "CD4",ident.2 = "CD14 Mono",only.pos = TRUE,verbose=F)
tmp <- retained[,retained$cellbender==TRUE]
m4 <- FindMarkers(tmp, ident.1 = "CD4",ident.2 = "CD14 Mono",only.pos = TRUE,verbose=F)
marker1 <- intersect(rownames(m1)[m1$p_val_adj<0.05],rownames(m2)[m2$p_val_adj<0.05])
marker2 <- intersect(rownames(m3)[m3$p_val_adj<0.05],rownames(m4)[m4$p_val_adj<0.05])
CD4.marker <- intersect(marker1,marker2)
df <- data.frame("logFC"=c(m1[CD4.marker,]$avg_logFC,
                           m2[CD4.marker,]$avg_logFC,
                           m3[CD4.marker,]$avg_logFC,
                           m4[CD4.marker,]$avg_logFC))
df$method <- c(rep("EmptyNN",length(CD4.marker)),rep("CellRanger 2.0",length(CD4.marker)),
              rep("EmptyDrops",length(CD4.marker)),rep("CellBender",length(CD4.marker)))
mu <- ddply(df, "method", summarise, grp.mean=mean(logFC))
ggplot(df, aes(x=logFC, color=method)) +
  geom_density()+xlab("Log2(foldchange)")+
  geom_vline(data=mu, aes(xintercept=grp.mean, color=method),linetype="dashed")
```


# Figure 2H
```{r include=TRUE}
label[,2] <- ifelse(label[,1]=="Negative","cell-free","cell-containing")
bcs <- rownames(nn.res$prediction)
rocobj1 <- roc(label[bcs,2], nn.res$prediction[bcs,'mean.crossval'])
rocobj2 <- roc(label[names(ranger.keep),2], as.numeric(ranger.keep))
bcs <- rownames(e.out[!is.na(e.out$FDR),])
rocobj3 <- roc(label[bcs,2], e.out[bcs,'FDR'])
rocobj4 <- roc(label[names(bender.keep),2], as.numeric(bender.keep))
ggroc(list("EmptyNN"=rocobj1,"CellRanger 2.0"=rocobj2,
           "EmptyDrops"=rocobj3,"CellBender"=rocobj4),
      legacy.axes = TRUE) + labs(x = "FPR", y = "TPR")+
  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 1), color="darkgrey", linetype="dashed")
```